%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../report"
%%% End:

Programming languages are the interface by which humans interact with machines. These languages comes in various shapes, forms and sizes, but the one thing they have in common is that they are all implemented through some sort of compiler. This report describes how we planned, designed and implemented a compiler, and thus created the Kite programming language. With that, the general aim of the project is to gain an understanding of compilers and programming languages, with a deepened comprehension of functional programming.

We start by presenting a general overview of compilers, properties of functional programming and an outline of the steps it takes to implement one. Thereafter, we specify our requirements of the language and compiler to be implemented. These will include fundamental features of pure functional languages and systems for ensuring the correctness of the written code.

In the first part of the design, we start to describe the language features of the Kite programming language. Here we show that the simplicity of functional programming is a very powerful concept, that can be utilized to write expressive and concise code. To exemplify, we use runnable snippets of code, written in Kite itself, to clearly highlight the purpose and value of each feature. We move on to outlining the different components of a compiler, and what specific role each component has. This will show how every component is invaluable for building a stable compiler.

Lastly, we describe in detail how we have chosen to implement the language. Here we use the Haskell programming language to step by step build a compiler from the ground up, utilizing the knowledge we have already gained through the planning process. Here it will be clear how a language like Haskell, is a good choice of language, since it already utilizes many of the features Kite will come to master. We also describe what further development on the compiler could lead to, and how we would have done it.

In conclusion, throughout the process of planning, designing and implementing the Kite programming language, we have gained a thorough understanding of compilers and functional programming languages. This is demonstrated by the functioning product that is Kite.
