%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../report"
%%% End:

We have been inspired by many other languages when designing and implementing Kite, most notably the Glasgow Haskell Compiler (GHC)\footnote{\url{http://www.haskell.org/ghc}}.


\subsection{Syntax and semantics}
In this section we will describe how to use the different features of Kite and how they work, i.e.\ the syntax and semantics of the language.

\subsubsection{Basic types and declarations}
Basic declaration of the variables in Kite, types are inferred by the analyzer:

\begin{kite}
one = 1
two = 2.0
truth = True
list = [1, 2, 3, 4]
char = 'a'
pair = (1, 'a')
\end{kite}

Functional languages often make extensive use of lists, which is indeed also the case for Kite. A list is an ordered array of items that can be transformed in many different ways. We use a common short-hand method for describing a list of things, using square brackets \code{[\ ]}. For instance, a \code{List(Int)} (pronounced ``List of Ints'') is denoted \code{[Int]} and a list of a type variables $a$ is \code{[a]}. Lists can be nested, allowing \code{List(List(Int))} denoted \code{[[Int]]}.

The pair type is also abbreviated using \code{( , )}, for instance the type \code{Pair(Int, Bool)} is written \code{(Int, Bool)}

Strings are represented as syntactic sugar for lists of characters. For
instance, \code{['f', 'o', 'o']} has the sugared version \code{"foo"}. This will be discussed later.

Basic arithmetic operators:
\begin{kite}
1 + 1  ----> 2
10 % 2 ----> 0
\end{kite}

\begin{table}[H]
  \centering
  \begin{tabular}{ll}
    \textbf{Operation} & \textbf{Meaning}   \\ \hline
    $\code{x + y}$     & Addition         \\ \hline
    $\code{x - y}$     & Subtraction        \\ \hline
    $\code{x * y}$     & Multiplication     \\ \hline
    $\code{x / y}$     & Division           \\ \hline
    $\code{x**y}$      & Exponentiation     \\ \hline
    $\code{x \% m}$    & Remainder of x / y \\ \hline
  \end{tabular}
  \caption{Basic arithmetic operators in Kite}
\label{tbl:kite-design-arith-op}
\end{table}

Basic boolean operators:
\begin{kite}
1 < 2  ----> True
1 == 2 ----> False
\end{kite}

\begin{table}[H]
  \centering
  \begin{tabular}{ll}
    \textbf{Operation} & \textbf{Meaning} \\ \hline
    $\code{==}$ & Equal                   \\ \hline
    $\code{/=}$ & Not equal               \\ \hline
    $\code{<=}$ & Less than or equal than \\ \hline
    $\code{>=}$ & Greater than or equal   \\ \hline
    $\code{<}$  & Less than               \\ \hline
    $\code{>}$  & Greater than            \\ \hline
  \end{tabular}
  \caption{Basic boolean operators in Kite}
\label{tbl:kite-disign-bool-op}
\end{table}


\subsubsection{Function declaration} TODO: infix function decl.
\label{sec:func-decl}
Every function in Kite is declared as an anonymous function:

\begin{kite}
|param1, ..., paramN| -> {
  expr1
  ...
  exprN
}
\end{kite}

A function (lambda expression) is defined as a list of arguments and a block. A block is defined as a list of expressions optionally separated by \code{;}, where it is implicit that the last expression is returned as the result of the functions. A block introduces a new lexical scope, capturing its parent scope.

There are no named functions in Kite, but lambda expressions can be bound to identifers. A function is declared as follows:

\begin{kite}
name :: ArgType -> ... -> ReturnType
name = |arg1, ...| -> { ... }
\end{kite}

Example of a valid function declaration in Kite, taking one parameter of the type \code{Int} and returning an \code{Int}:

\begin{kite}
foo :: Int -> Int
foo = |a| -> {
  a + 1
}
\end{kite}

A runnable program must always include the \code{main} function of type \code{Void -> a}, which will be used as the entry point for execution:

\begin{kite}
main = |Void| -> { ... }
\end{kite}

In order to use the preprocessor to include another code file, use the \code{\#include} keyword:
\begin{kite}
#include "foo.kite"
\end{kite}

\subsubsection{Type declaration}

Top-level declaration can optionally have a corresponding type declared. This can greatly increase readability of declarations as the inferred type might not be immediately clear.

Annotations are declared using the '\code{::}' operator. For instance the function \code{length} can have its type explicitly annotated as follows:

\begin{kite}
length :: [a] -> Int
length = |xs| -> {
  ...
}
\end{kite}

The type-checker will validate that an annotated type is at least as general as the inferred type. This means that a type annotation of \code{a} will match anything, and \code{a -> a} will match \code{Int -> Int} but not \code{Int -> Bool}

Types can be grouped to denote a higher-order function, for instance the type of map \code{(a -> b) -> [a] -> [b]} means that \code{map} accepts a \emph{function} as its first parameter, a list as its second and it produces a list.


\subsection{Function application}
TODO: partial app


\subsubsection{Syntactic sugar}
\label{sec:kite-design-sugar}
By using a sugaring module, Kite will be able to express certain functions more clearly and readable to the user. Syntactic sugar includes infix operators, multiple parameters on function calls, list of characters as strings and list comprehensions. Infix operators will allow infix use of binary functions (taking two parameters) in Kite. An example of infix operators is the basic $+$ arithmetic operation:

\begin{kite}
foo = 1 + 2
\end{kite}

Where the desugared version of the same expression is given as:

\begin{kite}
foo = ((+) (1)) (2)
\end{kite}

As Kite functions will only take one parameter, sugaring can simulate the use of multiple parameters in functions. Example hereof:

\begin{kite}
foo :: a -> a -> a
foo = |a, b| -> {
  a + b
}
\end{kite}

Which will be syntactic sugar for the following:

\begin{kite}
foo :: a -> a -> a
foo = |a| -> {
  return |b| -> {
    return ((+) (a)) (b)
  }
}
\end{kite}

\textbf{Parameterless functions} are also syntactic sugar. For instance the main function is often declared as follows:

\begin{kite}
main = -> {
  ...
}
\end{kite}

This is simply shorthand for

\begin{kite}
main = |Void| -> {
  ...
}
\end{kite}

As mentioned earlier, \textbf{Strings} are syntactic sugar for list of characters, \code{[Char]}, which will allow for easier representations of character lists. An example of a string:

\begin{kite}
str = "Hello"
\end{kite}

Where the corresponding desugar will be:
\begin{kite}
str = ['H', 'e', 'l', 'l', 'o']
\end{kite}
Note the distinction between strings using double-quotes ("...") and characters using single-quotes ('...').

\label{sec:ex-listcomp}
\textbf{List comprehensions} are syntactic sugar for manipulating lists. An example of a list comprehension is:

TODO: Mentiod output expr, draws 'n' grdz

\begin{kite}
list = [ x*y | x <- range(3,5), y <- range(4,6) | (x+y) >= 10, y > 4]
\end{kite}

Where the desugared version of the given example, will look as follows:

\begin{kite}
list =
flatMap(|x| -> {
  flatMap (|y| -> {
    if (|x,y| -> {y > 4})(x,y) && (|x,y| -> {(x+y) >= 10})(x,y)
       then [(|x,y| -> {x*y})(x,y)]
       else []
  } , range(4,6) )
}, range(3,5) )
\end{kite}

Note that list comprehensions uses functions defined in section~\ref{foundation}, which will be discussed later.

\textbf{Infix functions} Kite allows for use of infix functions. Infix functions are syntactic sugar for prefix function calls, with some given parameters. Using '\code{\`}' (backtick) will imply the use of infix function calls

A common example of infix function calls, is the \code{elem}-function, which is defined as follows (from \nameref{foundation}):
\begin{kite}
elem :: a -> [a] -> Bool
elem = |el| -> {
  any ((el==))
}
\end{kite}

The following Kite example will be an example of usage of \code{elem}. \code{elem} is here used as an infix function call.
\begin{kite}
list = [1, 2, 3]
1 `elem list ----> True
\end{kite}
The resulting desugared version of same example will look as follows:

\begin{kite}
list = [1,2,3];
((elem) (1)) (list); ----> True
\end{kite}

\subsubsection{Recursion}
Kite supports recursion, which allows functions to call themselves in programs. Basic recursion will have the following syntax:
\begin{kite}
foo :: Int -> Int
foo = |a| -> {
  if a == 1 then 1
  else a + foo(a - 1)
}
\end{kite}

Kite also supports recursion in pattern matching, which will be discussed next.

\subsubsection{Conditionals}
\label{subsubsec:conditionals}
Kite features two conditional expressions; the if-expression and pattern matching.

The \textbf{if-expression} is not to be confused with an if-statement, as our implementation is not a control flow statement. It will always return an expression, and the return-types of the consequent and alternative must be the same.

\begin{kite}
if booleanExpr then consequenceExpr else alternativeExpr
\end{kite}

A more advanced form of conditional is \textbf{pattern matching}. It works by matching an expression to a list of cases, which can be either a normal expression or it can be a destructuring\footnote{Destructuring means breaking up a data structure into its parts by using pattern matching. Kite features a simple implementation of this in its \code{match} expressions} of a list or a pair.

Below we show the matching of \code{n} on 0, which is an expression, and \code{otherwise}, which is written \code{"\_"}:

\begin{kite}
fibo :: a -> b -> Int -> a
fibo = |a, b, n| -> {
  match n {
    0 -> a,
    _ -> fibo(b, (a + b), (n - 1))
  }
}
\end{kite}

An example of pattern matching on lists is our implementation of the function \code{reverse} (\ref{foundation}), which matches \code{xs} with an expression, empty list, and the head and tail of \code{xs}:

\begin{kite}
reverse = |xs| -> {
  match xs {
    []    -> [],
    x,xs  -> reverse (xs) ++ [x]
  }
}
\end{kite}

The final case is matching on pairs. The example below shows the implementation of \code{fst} (first) in Foundation (\ref{foundation}):
\begin{kite}
fst :: (a, b) -> a
fst = |p| -> {
  match p {
    (a , b) -> a
  }
}
\end{kite}

\subsubsection{Higher-order functions}
Kite features higher-order functions (HoF), and the \code{map} function is an example hereof. It takes a function as argument and applies it to every element of a list, and returns the resulting list:

\begin{kite}
map :: (a -> b) -> [a] -> [b]
map = |f, ls| -> {
  match ls {
    [] -> [],
    x,xs -> f (x) : map (f, xs)
  }
}
\end{kite}

\subsubsection{Currying}
\label{subsubsec:currying}
Beside every function in Kite being curried (which should be clear from the type-declarations), we have also implemented a \code{curry} (and \code{uncurry}) function, which given a function with a single pair as the argument, returns the same function taking two arguments corresponding to the values of the pair:

\begin{kite}
curry :: ((a, b) -> c) -> (a -> b -> c)
curry = |f| -> {
  |a,b| -> {
    f((a,b))
  }
}
\end{kite}

Uncurry, naturally does the opposite, meaning it takes a function taking two parameters and returns the same taking a single pair as an argument.

\begin{kite}
uncurry :: (a -> b -> c) -> ((a, b) -> c)
uncurry = |f| -> {
  |p| -> {
    match p {
      (a, b) -> f(a,b)
    }
  }
}
\end{kite}

We have chosen to include these two functions from~\nameref{foundation} in this section as they reflect an essential part of Kite and functional programming languages in general.

\subsubsection{Partial application}
An example of partial application is the function \code{increment}, which partially applies addition:

\begin{kite}
increment :: Int -> Int
increment = (1+)
\end{kite}

Increment can then be applied to an integer:

\begin{kite}
increment(1) ----> 2
\end{kite}

It should be noted that the addition operator, \code{+}, has type \code{a
  -> a -> a}. Thus, partial application yields \code{a -> a} and since
it is partially applied with an integer, the type of \code{increment}
becomes \code{Int -> Int}

\subsubsection{Error handling}
Kite supports termination of execution using the function \code{panic}. Using \code{panic} will stop the execution. It has the type signature \code{String -> a}, thus being allowed anywhere.

An example of the \code{panic} function in use:
\begin{kite}
head :: [a] -> a
head = |xs| -> {
  match xs {
    [] -> panic ("can't take head of empty list"),
    x,xs -> x
  }
}
\end{kite}
If a user tries to take the head of an empty list, the execution of Kite will stop and the error message will be outputted.


\subsection{Foundation}
As mentioned earlier in section~\ref{sec:requirements}, the \code{Foundation} of Kite is a standard library implemented in Kite consisting of often used functions. Kite's Foundation is inspired by Haskell's \code{Prelude}, and includes many functions found in
Prelude.

Below are two examples from Foundation.

\subsubsection{List concatenation}
List concatenation does as the name suggests, namely concatenating two lists into a single list:

\begin{kite}
{++} :: [a] -> [a] -> [a]
{++} = |l, r| -> {
  match l {
    [] -> r,
    x,xs -> x : (xs ++ r)
  }
}
\end{kite}

For example, as strings are simply lists of characters, \code{++} can be used as follows:

\begin{kite}
"Hello," ++ " World!" ----> "Hello, World!"
\end{kite}

\subsubsection{List indexing}
Lists in Kite are zero-indexed, meaning that the first element of a list is at 'index' zero. Indexing a list retrieves the $n$-th element of a list:

\begin{kite}
{#} :: [a] -> Int -> a
{#} = |xs, idx| -> {
  match xs {
    [] -> panic ("list index too large"),
    x,xs -> if idx == 0 then x else xs # (idx - 1)
  }
}
\end{kite}

A simple example of list indexing is as follows:

\begin{kite}
teens = range(13,19)
sweetSixteen = teens # 3 ----> 16
\end{kite}

For a full documentation of Foundation, please see section~\ref{foundation}.
